**The Heart Lies: The Role of Interoception and Theory of Mind in Deception**

### 2022-11-04 paper/manuscript.Rmd
update manuuuu

In our current post-truth era, deception has become increasingly prevalent in our everyday lives [@lewandowsky2017beyond]. Although the act of deceiving is complex and cognitively taxing [@abe2007deceiving], often requiring effortful self-regulation to conceal [@krokoszinski2016emotion], research shows that people tend to under-estimate their ability to successfully deceive others [@elaad2003effects]. Moreover, meta-analytic findings suggest that the ability of the liar plays a more significant role than the lie detector in deceptive interactions [@verigin2019lie; @bond2008individual; @levine2011sender]. However, relative to the abundant literature regarding deception detection [@viji2022history; @sternglanz2019review; @masip2017deception], surprisingly little attention has been conferred to investigating the inter-individual differences in people's ability to successfully lie.

@@

As deception requires the liar to intentionally manipulate the beliefs of others [@burgoon1994interpersonal; @sip2012if], most research on lying capability has been focused on higher order cognitive processes, particularly in theory of mind (TOM). TOM refers to the ability to infer that others have mental states, such as beliefs, emotions and intentions, distinct from ourselves [@lee2021lying; @wellman2001meta; @baron1997mindblindness]. Indeed, the capacity to tell lies, as well as the complexity of lies told, have previously been found to be associated with higher TOM abilities [@talwar2007lying; @evans2011verbal; @talwar2017role]. Furthermore, the deceptive capabilities of children has often been used as a measure of their TOM abilities. However, given that the development of TOM abilities has often been viewed as the building blocks of other higher order cognitive processes [@ketelaars2010dynamics; @isaksson2021s], studies investigating the links between TOM and lying ability have largely been conducted with children and individuals with neurocognitive deficits [@beaudoin2020systematic; @bora2017meta; @roheger2022progression]. As such, little is known with regards to the role TOM plays in the lying ability of neurotypical adults.

@@

In contrast, however, the somatic marker hypothesis posits higher interoception to be linked to a propensity towards making rational decisions [@damasio1996somatic]. Indeed, previous research has shown a negative correlation between interoceptive awareness and the likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. Taking these mixed findings in perspective with the relative lack of deception research associated with the multidimensional construct of interoception, more studies are warranted to investigate the relationship between the interoceptive and deceptive ability.

@@

<!-- This research attempts to isolate the effects of theory of mind and interoception to examine how participants may rely on these abilities to evade lie detection. Of particular interest is how liars might utilize their interoceptive ability to self regulate in towards a more successful deception, and how this ability may compare or interact with other mechanisms at play such as theory of mind. We hypothesize that participants with greater theory of mind ability will show higher lie confidence, shorter reaction time, and lower physiological arousal in the interrogation condition, while participants in the polygraph condition will demonstrate lower lie confidence, lower reaction time and higher physiological arousal. -->

@@

Along with open science standards, the material (stimuli generation code, experiment code, raw data, analysis script with complementary figures and analyses, preregistration, etc.) for this research is available at: **[LINK].**

@@

*26* participants (65.4% Females, M~*age*~ = 20.9, *SD* = 2.0) were recruited advertising using university mailing lists and posters. Participants were awarded with academic credit for their participation. This research was approved by the **Institutional Review Board (Reference Number: IRB-)** of Nanyang Technological University (NTU).

@@

Since theory of mind and empathy are closely related, *The Basic Empathy Scale (BES)* [@jolliffe2006development] was used as a subjective measure for theory of mind. *BES* uses a 40-item Likert scale to rate statements such as, "I can usually work out when people are cheerful" from 1-5. We also implemented the *Yoni Task,* which assesses the ability to judge mental states based on verbal cues, eye gaze and facial expression [@shamay2007dissociable].

@@

To measure interoception, participants completed a *Heartbeat Counting Task (HCT)* [@schandry1981heart], where participants count their heart beats without taking their pulse, as well as provide a confidence rating for their estimate. At the same time their true heart rate is recorded and the scores are compared to determine the individual's accuracy. They also completed the *MAIA questionnaire* [@mehling2012multidimensional], that featured 32 statements such as "I can use my breath to reduce tension" on a 1-5 Likert scale.

### Deception

We used the *LIE Scale*, which is a 44-item questionnaire examining deception as a dispositional trait, and measures the participant propensity towards lie frequency, ability, contextuality and negativity [@makowski2021structure].

@@

The demographic information we collected was age, ethnicity, nationality, dominant hand (for application of EDA electrodes), education level (determined by highest qualification received, or in the midst of achieving), religion and duration lived in Singapore.

@@

Each subject was briefed on the experimental procedure and study aims, and signed an informed consent document prior to the study commencing. During the first phase of the experiment, demographic measures were taken, followed by measures of Theory of Mind and interoceptive abilities, as described above. Participants were then connected to an ECG before proceeding with the deception task on a computer.

#### Deception Task

Participants underwent 40 trails of a directed lying task, where they were asked questions such as, "What secondary school did you go to?". Answers were provided verbally and participant pressed a button when they were finished, at which point they would be connected for 10 seconds to the condition-unique stimulus. There were two conditions of 20 trails each, beginning with the *interrogation condition*, and followed by the second, *polygraph condition*. The purpose of these conditions was to draw out the interoceptive or theory of mind abilities by providing feedback cues that would cause them to focus their attention on either the visual feedback cues from other people, or their own cardiac response.

In the *interrogation condition*, participants were instructed that following their lie, a live examiner over video feed would appear briefly to judge whether they were truthful or not. In actuality, the video of the examiner shown was pre-recorded and the same for all participants. We hypothesized that participants would will rely on theory of mind abilities to predict whether their lie to the examiner was successfully accepted as true. We hypothesized that in this condition, participants would rely on their theory of mind abilities to discern the facial feedback cues from the examiner to ascertain whether their lie was successful.

The *polygraph condition* followed the same routine, however instead of a receiving feedback from an interrogator, following the lie, a video of physiological signals was shown. We hypothesized that when participants were made acutely aware of such signals through false feedback, they would rely on interoceptive abilities to be more mindful or intentional about modulating the signals that may give them away. The physiological signals in this condition are also false feedback, and pre-recorded and identical for all participants.

During each trial, the time it took for participants to answer was recorded, and their heart rates are measured using an ECG. Following the trial, participants gave a confidence rating for how successful they believe their lie or truth was. Debriefing was provided following the study. At this time the truthful answers to the directed lie questions were collected via questionnaire.

@@

The manipulation checks consist of testing whether there is an effect of question phrasing (direct vs. indirect), and condition (polygraph *vs.* interrogation), on our 3 outcome variables: the participants' ratings of confidence that their answers (lies *vs.* truths) were convincing, the response time (RT), and the heart rate change associated with the response. This analysis was performed using mixed models with the participants and questions both entered as random factors. Marginal contrasts analysis was also performed to clarify the differences between conditions. To allow for a better quantification of the uncertainty associated with the effects, as well increase the robustness to outliers and artifactual findings, all statistics were undertaken under the Bayesian framework [**ref makowski2019existence**], using informative priors centred around 0 ($t_{Confidence}(1, 0, 1)$, $t_{RT}(1, 0, 3)$, $t_{Heartrate}(1, 0, 8)$).

@@

The data analysis was carried out using R, the *brms* package [*REF*] and the *easystats* ecosystem **[REFS]**. As the full reproducible analysis script and statistical results are available at **[INSERT LINK]**, we will only focus on significant results in the manuscript.

@@

The 3 YONI-task dimensions and the 2 BES traits were combined into a unique factor, labelled *Theory of Mind* (explaining 35.76% of variance). It was loaded by the cognitive (.89), affective (.77), physical (.45) YONI dimension, and the affective (.41) and cognitive (.17) traits of the BES.

@@

The composite *ToM* score was significantly associated with less confident ($\beta=-0.19,~95\%~CI~[-0.36, -0.02],~pd = 98.47\%$) and slower lies ($\beta=0.42,~95\%~CI~[0.01, 0.83],~pd = 97.67%\%$), specifically in the polygraph condition. No significant effect was found heart rate, and no correlation with the deception scale traits.

>>>>

In this post-truth era, deception has become increasingly prevalent in our everyday lives [@lewandowsky2017beyond]. Although the act of deceiving is complex and cognitively taxing [@abe2007deceiving], often requiring effortful self-regulation to conceal [@krokoszinski2016emotion], research shows that people tend to under-estimate their ability to successfully deceive others [@elaad2003effects]. Moreover, meta-analytic findings suggest that the ability of the liar plays a more significant role than the lie detector in deceptive interactions [@verigin2019lie; @bond2008individual; @levine2011sender]. However, relative to the abundant literature regarding deception detection [@viji2022history; @sternglanz2019review; @masip2017deception], surprisingly little attention has been conferred to investigating the inter-individual differences in people's ability to successfully lie.

@@

As deception requires the liar to intentionally manipulate the beliefs of others [@burgoon1994interpersonal; @sip2012if], most research on lying capability has been focused on higher order cognitive processes, particularly in theory of mind (TOM). TOM refers to the ability to infer that others have mental states, such as beliefs, emotions and intentions, distinct from ourselves [@lee2021lying; @wellman2001meta; @baron1997mindblindness]. Indeed, the capacity to tell lies, as well as the complexity of lies told, have previously been found to be associated with higher TOM abilities [@talwar2007lying; @evans2011verbal; @talwar2017role]. Furthermore, the deceptive capabilities of children is often used as a measure of their TOM abilities. However, given the view that development of TOM abilities serve as the building blocks of other higher order cognitive processes [@ketelaars2010dynamics; @isaksson2021s], studies investigating the links between TOM and lying ability have predominantly been focused on children and individuals with neurocognitive deficits [@beaudoin2020systematic; @bora2017meta; @roheger2022progression]. As such, little is known with regards to the role TOM plays in the lying ability of neurotypical adults.

@@

In contrast, however, the somatic marker hypothesis posits higher interoception to be associated with a propensity towards making rational decisions [@damasio1996somatic]. Indeed, previous research has shown a negative correlation between interoceptive awareness and the likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. Taking these mixed findings to perspective with the lack of consideration for the multidimensional nature of interoception in past studies, as well as the paucity in research linking deceptive ability with interoception, more studies are warranted to examine the relationship between interoception and deception.

@@

@@

Along with open science standards, the material (stimuli generation code, experiment code, raw data, analysis script with complementary figures and analyses, preregistration, etc.) for this research is available at [**https://github.com/DominiqueMakowski/DeceptionInteroTom**](https://github.com/DominiqueMakowski/DeceptionInteroTom){uri.}.

@@

30 participants were recruited through  posters, flyers, and online social media platforms. Inclusion criteria included residing in Singapore, and an absence of current or history of neurological, psychiatric, or cardiovascular disorders. 4 participants were excluded as a result of a technical error that led to their data not being recorded. The final sample consists 26 participants (Mean age = 20.9, SD = 2.0, range:[18, 25], Sex: 65.4% women, 34.6% men, 0.00% non-binary). The heart rate of one participant and response time of one participant were excluded from analysis due to extreme outlying values [**PLEASE HELP ME CONFIRM**].

This study was approved by the NTU Institutional Review Board (NTU IRB-2018-10-029)(**TBC**) and all procedures performed were in accordance with the ethical standards of the institutional board and with the 1964 Helsinki Declaration. All participants provided their informed consent prior to participation and were awarded with academic credit upon completion of the study.

@@

The *Basic Empathy Scale (BES)* is a 20-item, 5-point Likert scale (Strongly Disagree - Stongly Agree) that measures cognitive and affective empathy [@jolliffe2006development]. Although TOM and empathy are regarded as distinct psychological constructs, abundant evidence points to them being closely related [@gallant2020developing; @sebastian2012neural], with empathy thought to be integrated in the affective dimension of TOM (i.e., the ability to infer what someone else is feeling) [@shamay2010role]. The *Yoni Task*, which comprises 3 dimensions, namely cognitive, physical and affective, was also administered to assess participants' objective TOM ability [@shamay2007dissociable].

@@

To measure interoceptive accuracy, participants completed a *Heartbeat Counting Task (HCT)* [@schandry1981heart] while having their heart beats recorded. During the HCT task, participants were instructed to count the number of heart beats they felt over 5 varying time periods (20s, 25s, 30s, 35s, 40s) administered in a randomized order. After each time period, interoceptive sensibility was assessed by having participants respond how confident they were of their counting accuracy using a visual analog scale. Furthermore, participants also completed the *MAIA-2 questionnaire* [@mehling2012multidimensional], a 32-item scale that measures 8 dimensions of interoception (Noticing, Not-Distracting, Not-Worrying, Attention, Regulation, Emotional Awareness, Self-Regulation, Body Listening, and Trust) using a 5-point Likert scale(Never-Always).

### Lie Profile

We used the *LIE Scale*, a 16-item questionnaire that assesses 4 dispositional lying dimensions (Ability, Negativity, Contextuality and Frequency) on a visual analog scale (Disagree-Agree) [@makowski2021structure].

@@

We further collected demographic information including age, ethnicity, nationality, handedness (for application of EDA electrodes), religion, duration lived in Singapore as well as highest education qualification achieved (or in the midst of achieving).

### Deception Task Paradigm

A novel and computerized directed-lying experimental paradigm was designed to assess participants' lying ability in different scenarios. Each *Deception Task* trial is comprised of a question presented on a computer screen with an accompanying instruction that either asked participants to verbally lie or tell the truth. The instructions for each trial were displayed briefly, for a maximum of 10 seconds, or until the participants responded accordingly. Notably, these questions were either phrased directly (*"What is your favourite sport?*) or indirectly (*"Is your favourite sport Hockey?"*). The *Deception Task* paradigm consists of 2 conditions - namely, the *Interrogation Condition* and the *Polygraph Condition*.

In the **Interrogation Condition**, social feedback in the form of a video stream with an examiner was presented to participants during each trial. Participants were informed that the examiner would be evaluating the truthfulness of their responses by observing through the live video feed, when in actuality, the video was a pre-recorded video clip of a confederate. This was done to heighten the salience of social cues so as to facilitate participants' TOM abilities.

In the *Polygraph condition*, cardiac feedback in the form of an electrocardiogram (ECG) was presented to participants during each trial Participants were informed that they were being shown live feedback of their own physiological signals when in fact, the displayed ECG was of a prerecorded video clip. This was done in order to heighten the salience of interoceptive cues and facilitate participants' interoceptive abilities.

Following each response, participants were asked to rate their confidence in the success of their truth or lie-telling before progressing to the next trial.

@@

In the first phase of the study, participants answered a brief demographic survey before being administered a battery of questionnaires (*BES*, *MAIA-2*, *Lie Scale*) and cognitive-behavioural tasks (*HCT*, *Yoni Task*). This series of tasks and questionnaires were administered in a randomized order [**TBC**] in order to ensure there were no order effects. In the second phase of the study, the *Deception Task* paradigm was presented to participants. Similar to the first phase, the order sequence of the 2 conditions (*polygraph* vs *interrogation*) was randomized [**TBC**], with each condition comprising 20 trials. In total, each participant underwent 40 *Deception Task* trials, of which they had to respond with either lies or truths depending on the given instruction.

Physiological recording devices were prepared prior to beginning the experimental paradigm. Using three ECG electrodes placed according to a modified Lead II configuration [@takuma1995alternative], cardiovascular activity was recorded via the BioPac MP160 system (BioPac Systems Inc., USA) and the AcqKnowledge 5.0 software. Events were marked by a photosensor attached to the screen.

@@

The manipulation checks consist of testing whether there is an effect of question phrasing (direct vs. indirect), and condition (polygraph *vs.* interrogation), on our 3 outcome variables: the participants' ratings of confidence that their answers (lies *vs.* truths) were convincing, the response time (RT), and the heart rate change associated with the response. This analysis was performed using mixed models with the participants and questions both entered as random factors. Marginal contrasts analysis was also performed to clarify the differences between conditions. To allow for a better quantification of the uncertainty associated with the effects, as well increase the robustness to outliers and artifactual findings, all statistics were undertaken under the Bayesian framework [@makowski2019bayestestr], using informative priors centred around 0 ($t_{Confidence}(1, 0, 1)$, $t_{RT}(1, 0, 3)$, $t_{Heartrate}(1, 0, 8)$).

@@

The data analysis was carried out using*R 4.2* [@RCoreTeam2022], *brms* [@Burkner2017], and the *easystats* collection of packages [@bayestestRArticle; @correlationArticle; @performanceArticle; @insightArticle]. As the full reproducible analysis script and statistical results are available at [**https://github.com/DominiqueMakowski/DeceptionInteroTom**](https://github.com/DominiqueMakowski/DeceptionInteroTom){.uri}, we will only focus on significant results in the manuscript.

@@

The 3 YONI-task dimensions and the 2 BES traits were combined into a unique factor, labelled *TOM* (explaining 35.76% of variance). It was loaded by the cognitive (.89), affective (.77), physical (.45) YONI dimension, and the affective (.41) and cognitive (.17) traits of the BES.

@@

The composite *TOM* score was significantly associated with less confident ($\beta=-0.19,~95\%~CI~[-0.36, -0.02],~pd = 98.47\%$) and slower lies ($\beta=0.42,~95\%~CI~[0.01, 0.83],~pd = 97.67%\%$), specifically in the polygraph condition. No significant effect was found heart rate, and no correlation with the deception scale traits.
